Efficient Machine Learning Approaches for Fast Risk Evaluation of VAs

### 2025-03-17 pre_private/pre.tex
1

$$\rho(L) = \rho(L(X)), \;\;\; L(X) = \mathbb{E}\left[ Y|X=x \right]\vert_{x=X}  $$

Involves two levels of Monte Carlo simulations:
\begin{itemize}
    \item Outer level: generates underlying risk factors (outer scenarios), $X_i \sim F_X$
    \item Inner level: generates scenario-wise samples of portfolio losses (inner replications), $Y_{ij} \sim F_{Y|X_i}$
\end{itemize}

@@

@@

\begin{itemize}
    \item Smooth $h$, e.g., quadratic tracking error

    $$ \rho(L) = \mathbb{E} \left[ (L - b)^2 \right] $$

    \item hockey-stick $h$: mean excess loss

    $$ \rho(L) = \mathbb{E} \left[ L \cdot \mathbbm{1}_{\{L \geq u\}} \right] $$

    \item indicator $h$: probability of large loss

    $$ \rho(L) = \mathbb{E} \left[ \mathbbm{1}_{\{L \geq u\}} \right] $$

    \item Value at Risk (VaR)

    $$ \rho_\alpha(L) = Q_\alpha(L) = \inf \{ u: \mathbb{P}(L \leq u) \geq \alpha \} $$

    \item Conditional Value at Risk (CVaR) \footnotemark

    $$ \rho_\alpha(L) = \mathbb{E} \left[ L \vert L \geq Q_\alpha(L) \right] $$

@@

@@

@@

@@

@@

@@

@@

\begin{frame}{Learning Lapse Features}
    \begin{figure}[ht!]
        \centering
        \begin{subfigure}{0.48\textwidth}
            \includegraphics[width=\textwidth]{../project3/figures/figure1a.png}
            \caption{Extensive Training on Target Task}
        \end{subfigure}
        \begin{subfigure}{0.48\textwidth}
            \includegraphics[width=\textwidth]{../project3/figures/figure1b.png}
            \caption{Without TL}
        \end{subfigure}
        \caption{Direct training on RS-GBM GMMB with static lapse}
    \end{figure}
\end{frame}

\begin{frame}{Learning Lapse Features}
    \begin{figure}[ht!]
        \centering
        \begin{subfigure}{0.48\textwidth}
            \includegraphics[width=\textwidth]{../project3/figures/figure1c.png}
            \caption{With Fine-tuning}
        \end{subfigure}
        \begin{subfigure}{0.48\textwidth}
            \includegraphics[width=\textwidth]{../project3/figures/figure1d.png}
            \caption{With Layer Freezing}
        \end{subfigure}
        \caption{TL on RS-GBM GMMB with static lapse}
    \end{figure}

@@

>>>>

$$\rho(L) = \rho(L(X)), \;\;\; L(X) = \mathbb{E}\left[ Y|X=x \right]\vert_{x=X}.  $$

Involves two levels of Monte Carlo simulations:
\begin{itemize}
    \item Outer level: generates underlying risk factors (outer scenarios), $X_i \sim F_X$.
    \item Inner level: generates scenario-wise samples of portfolio losses (inner replications), $Y_{ij} \sim F_{Y|X_i}$.
    \item Total simulation budget: $\Gamma = M \cdot N$.
\end{itemize}

@@

\note{In the context of quantitative risk management, nested simulation is used for risk evaluation of financial derivatives and insurance products.}
\note{The nested structure is due to the fact that the loss random variable is in the form of a conditional expectation of the portfolio loss given the underlying risk factors.}
\note{In our problems, the loss random variable $L$ depends on the risk factors $X$ and is the conditional expectation of the portfolio loss given $X$.}
\note{Estimating a risk measure, $\rho(L)$, involves two levels of Monte Carlo simulations: outer level generates underlying risk factors, and inner level generates scenario-wise samples of portfolio losses.}
\note{The nested structure makes the estimation computationally expensive.}
\note{In this thesis, we focus on examining and improving the efficiency of nested simulation procedures mostly using supervised learning metamodels.}

@@

\begin{itemize}
    \item Smooth $h$, e.g., quadratic tracking error,

    $$ \rho(L) = \mathbb{E} \left[ (L - b)^2 \right]; $$

    \item hockey-stick $h$: mean excess loss,

    $$ \rho(L) = \mathbb{E} \left[ L \cdot \mathbbm{1}_{\{L \geq u\}} \right]; $$

    \item indicator $h$: probability of large loss,

    $$ \rho(L) = \mathbb{E} \left[ \mathbbm{1}_{\{L \geq u\}} \right]; $$

    \item Value at Risk (VaR),

    $$ \rho_\alpha(L) = Q_\alpha(L) = \inf \{ u: \mathbb{P}(L \leq u) \geq \alpha \}; $$

    \item Conditional Value at Risk (CVaR) \footnotemark,

    $$ \rho_\alpha(L) = \mathbb{E} \left[ L \vert L \geq Q_\alpha(L) \right]. $$

@@

\note{We consider different types of risk measures in this thesis.}
\note{The first three are in expectation form, and the last two are quantile-based risk measures.}
\note{The difference between the first three is the choice of the function $h$.}
\note{$h$ can be a smooth function, a hockey-stick function, or an indicator function.}
\note{The examples of risk measures are quadratic tracking error, mean excess loss, and probability of large loss.}
\note{The last two are value at risk and conditional value at risk, which are widely used in practice.}
\note{Note that the definition of CVaR requires the knowledge of the Value at Risk.}

@@

\note{In our first project, we focus on comparing existing nested simulation procedures for different types of risk measures.}
\note{They differ in the way they estimate $L(X_i)$, the portfolio loss conditional on the risk factors $X_i$.}
\note{The standard nested simulation procedure uses the inner sample mean under that scenarioto estimate $L(X_i)$.}
\note{The optimal growth order of $M$ and $N$ is found by \citet{gordy2010nested}.}
\note{For a growing total simulation budget $\Gamma$, $M$ should grow in the order of $\Gamma^{2/3}$, and $N$ should grow in the order of $\Gamma^{1/3}$.}
\note{For a finite-sample case, the bootstrap method is used to estimate the optimal $M$ and $N$ by \citet{zhang2021bootstrap}.}
\note{Since the standard procedure only uses the information from the same scenario, it is potentially \textbf{wasteful} use of budget.}
\note{Subsequent works focus on improving the efficiency of nested simulation by pooling inner replications from different outer scenarios.}

@@

\note{Pooling can happen in different ways, and the key idea is to use the information from different outer scenarios to improve the efficiency of nested simulation.}
\note{There are two main streams of research in this direction.}
\note{The first stream is to use a supervised learning metamodel to approximate the simulation.}
\note{A metamodel is a model of model, or a surrogate model for the simulation model.}
\note{Here, the metamodel is used to approximate the inner simulation model, by treating it as a black-box function.}
\note{The second stream is to use importance sampling to improve the efficiency of nested simulation.}
\note{The likelihood ratio weights are used to weight the inner replications from different outer scenarios.}
\note{This end up improving the efficiency of nested simulation as well.}

@@

\note{A supervised learning based nested simulation procedure is implemented following these steps.}
\note{First, the user runs the standard nested simulation procedure to generate a set of outer scenarios and inner replications.}
\note{Then, the outer scenarios and inner sample means can be treated as features and labels.}
\note{Their relationship is then approximated by a supervised learning metamodel.}
\note{Finally, the trained metamodel is used to make predictions for all $X \in \mathcal{X}$.}
\note{In other words, instead of using the inner sample mean to estimate $L(X_i)$, we use the trained metamodel to make predictions of the portfolio loss.}
\note{These loss predictions are then used to calculate the risk measures of interest.}
\note{For most research in this area, the focus is on the convergence rate of the risk measure estimator in terms of the simulation budget.}
\note{However, in this thesis, we note that there are other important aspects to consider, such as the cost of pooling that may come from metamodel training and validation.}

@@

\note{For a nested simulation procedure with a given total simulation budget $\Gamma$, the objective is to find the optimal $M$ and $N$ that minimizes the MSE of the estimator.}
\note{Thus, in this thesis, we compare the asymptotic convergence rates of different nested simulation procedures using their MSE.}
\note{Here, $\hat{\rho}_{M, N}$ is the risk measure estimator using the nested simulation procedure with $M$ outer scenarios and $N$ inner replications.}
\note{We also consider the cost of pooling that may come from metamodel training and validation in our numerical experiments.}

@@

\note{We start by comparing the asymptotic convergence rates of different nested simulation procedures.}
\note{}

@@

\begin{frame}{The Effect of Number of Training Samples}
    \begin{figure}[ht]
        \centering
        \begin{subfigure}{0.48\textwidth}
            \includegraphics[width=\textwidth]{../project3/figures/figure1a.png}
            \caption{50000 samples}
        \end{subfigure}
        \begin{subfigure}{0.48\textwidth}
            \includegraphics[width=\textwidth]{../project3/figures/figure1b.png}
            \caption{2000 samples}
        \end{subfigure}
        \caption{Learning curves of GMMB contracts with static lapse}
    \end{figure}

@@

\begin{frame}{Learning Lapse Features using Transfer Learning}

    \begin{figure}[ht]
        \centering
        \begin{subfigure}{0.48\textwidth}
            \includegraphics[width=\textwidth]{../project3/figures/figure1c.png}
            \caption{Fine-tuning}
        \end{subfigure}
        \begin{subfigure}{0.48\textwidth}
            \includegraphics[width=\textwidth]{../project3/figures/figure1d.png}
            \caption{Layer Freezing}
        \end{subfigure}
        \caption{Learning curves of GMMB contracts with static lapse}
    \end{figure}

\begin{frame}{Learning Curves: Learning Dynamic Lapse}

    \begin{figure}[ht]
        \centering
        \begin{subfigure}{0.48\textwidth}
            \includegraphics[width=\textwidth]{../project3/figures/figure2a.png}
            \caption{No Lapse}
        \end{subfigure}
        \begin{subfigure}{0.48\textwidth}
            \includegraphics[width=\textwidth]{../project3/figures/figure2b.png}
            \caption{Static Lapse}
        \end{subfigure}
        \caption{Learning curves of GMMB contracts with dynamic lapse}
    \end{figure}

    \begin{frame}{Results: Accuracy Comparison}
    \begin{table}
    \begin{tabular}{lcccc}
    \toprule
    \textbf{Lapse Type} & \textbf{Extensive} & \textbf{Fine-tuning} & \textbf{Layer Freezing} & \textbf{Without TL} \\
    \midrule
    No Lapse & N/A & 0.4894 & \textbf{0.3361} & N/A \\
    Static Lapse & N/A & 0.0794 & 0.0763 & N/A \\
    Dynamic Lapse & \textbf{0.0587} & N/A & N/A & \textbf{0.2950} \\
    \bottomrule
    \end{tabular}
    \end{table}

    \textit{Comparison of different TL methods on GMMB contracts (best MSE values)}
    \end{frame}

    \begin{frame}{Learning Curves: Effect of Similarity}
    \begin{columns}
    \column{0.5\textwidth}
    \includegraphics[height=0.8\textheight]{../project3/figures/figure3a.png}
    \column{0.5\textwidth}
    \includegraphics[height=0.8\textheight]{../project3/figures/figure3b.png}
    \end{columns}

    \textit{The effect of similarity between source and target on the convergence speed}
    \begin{itemize}
        \item \textit{Left: from No Lapse; right: from Static Lapse}
    \end{itemize}
    \end{frame}

    \begin{frame}{Transfer Knowledge to other Contract Types}
    \textbf{From GMMB to GMWB Contract}
    \end{frame}

    \begin{frame}{Learning Curves: Transfer Knowledge to other Contract Types}
    \begin{columns}
    \column{0.5\textwidth}
    \includegraphics[height=0.8\textheight]{../project3/figures/figure4a.png}
    \column{0.5\textwidth}
    \includegraphics[height=0.8\textheight]{../project3/figures/figure4b.png}
    \end{columns}

    \textit{Learning curves of GMWB contracts}
    \begin{itemize}
        \item \textit{Left: 50000 samples; right: 2000 samples}
    \end{itemize}
    \end{frame}

    \begin{frame}{Learning Curves: Effect of Similarity}
    \begin{columns}
    \column{0.5\textwidth}
    \includegraphics[height=0.8\textheight]{../project3/figures/figure4c.png}
    \column{0.5\textwidth}
    \includegraphics[height=0.8\textheight]{../project3/figures/figure4d.png}
    \end{columns}

    \textit{Comparison of different TL methods on GMWB contracts}
    \begin{itemize}
        \item \textit{Left: fine-tuning; right: layer freezing}
    \end{itemize}
    \end{frame}

    \begin{frame}{Performance Comparison (GMMB → GMWB)}
    \begin{table}
    \begin{tabular}{lcc}
    \toprule
    \textbf{Model} & \textbf{Training MSE} & \textbf{True MSE} \\
    \midrule
    Without TL & 0.3588 & 0.4188 \\
    Fine Tuning & 0.1690 & 0.1780 \\
    Layer Freezing & 0.1828 & 0.2295 \\
    Extensive Training & 0.0853 & 0.0726 \\
    \bottomrule
    \end{tabular}
    \end{table}

    \begin{itemize}
        \item Fine-tuning outperforms layer freezing for dissimilar tasks
        \item Both TL methods better than training from scratch
        \item Extensive training still superior with abundant data
    \end{itemize}
    \end{frame}

    \begin{frame}{Practical Implementation}
    \begin{itemize}
        \item \textbf{Framework}: TensorFlow/PyTorch implementation
        \item \textbf{Workflow}:
        \begin{enumerate}
            \item Train base model on source contract
            \item Freeze early layers
            \item Fine-tune later layers on target contract
            \item Deploy for production use
        \end{enumerate}
    \end{itemize}

    \textbf{PyTorch} is recommended.
    \begin{itemize}
        \item It is more flexible and easier to customize
        \item It is more intuitive and easier to understand 
        \item It is more efficient and easier to debug
    \end{itemize}
    \end{frame}

    \begin{frame}{Multi-task Learning}
    \includegraphics[width=0.9\textwidth]{../project3/figures/mtl.png}

    \begin{itemize}
        \item LSTM layers shared across multiple tasks
        \item Task-specific fully connected layers
        \item Objective: Minimize sum of loss functions across all tasks
    \end{itemize}
    \end{frame}

    \begin{frame}{Multi-task Learning Framework}
    \textbf{Input}: Set of $K$ tasks $\{\mathcal{T}_k\}_{k=1}^K$ with datasets $\mathcal{D}_k = \{(X_k^{(i)}, L_k^{(i)})\}_{i=1}^{M_k}$, shared parameters $\theta_0$, and task-specific parameters $\theta_k$ for each task $k$

    \textbf{Algorithm}:
    \begin{enumerate}
        \item Train the multi-head LSTM metamodel on all $K$ tasks simultaneously by minimizing the multi-task loss function:
        \begin{equation}
            \min_{\theta_0, \{\theta_k\}_{k=1}^K} \sum_{k=1}^K \frac{1}{M_k} \sum_{i=1}^{M_k} \left( f_i(X_k^{(i)}; \theta_0, \theta_k) - L_k^{(i)} \right)^2
        \end{equation}
        \item Update both the shared parameters $\theta_0$ and task-specific parameters $\{\theta_k\}_{k=1}^K$ simultaneously using backpropagation and gradient descent with learning rate $\alpha$
    \end{enumerate}

    \textbf{Output}: Trained multi-task LSTM metamodel $f(\cdot; \theta_0, \{\theta_k\}_{k=1}^K)$ for all $K$ tasks
    \end{frame}

    \begin{frame}{Multi-task Learning: GMMB and GMWB}
    \begin{columns}
    \column{0.5\textwidth}
    \includegraphics[height=0.8\textheight]{../project3/figures/figure5a.png}
    \column{0.5\textwidth}
    \includegraphics[height=0.8\textheight]{../project3/figures/figure5b.png}
    \end{columns}

    \textit{Multi-task learning of GMMB and GMWB}
    \begin{itemize}
        \item \textit{Left: total MSE; right: individual MSE}
    \end{itemize}
    \end{frame}

    \begin{frame}{Multi-task Learning: GMMB and GMWB}
    \begin{columns}
    \column{0.5\textwidth}
    \includegraphics[height=0.8\textheight]{../project3/figures/figure5c.png}
    \column{0.5\textwidth}
    \includegraphics[height=0.8\textheight]{../project3/figures/figure5d.png}
    \end{columns}

    \textit{Training without multi-task learning}
    \begin{itemize}
        \item \textit{Left: GMMB; right: GMWB}
    \end{itemize}
    \end{frame}

    \begin{frame}{Challenges \& Limitations}
    \begin{itemize}
        \item Determining optimal layer freezing strategy
        \item Handling significantly different contract structures
        \item Quantifying uncertainty in transfer learning predictions
        \item Regulatory acceptance of black-box approaches
    \end{itemize}
    \end{frame}

    \begin{frame}{Conclusions}
    \begin{itemize}
        \item Transfer learning significantly improves metamodeling for VA contracts:
        \begin{itemize}
            \item Faster training convergence
            \item Better prediction accuracy
            \item Reduced computational requirements
        \end{itemize}
        \item Enables more frequent risk assessments and faster decision-making
    \end{itemize}
    \end{frame}

    \begin{frame}{Future Directions}
    \begin{itemize}
        \item Incorporating domain knowledge into transfer process
        \item Extension to other insurance and financial products
        \item Multi-task learning with more than two tasks
    \end{itemize}
    \end{frame}
